---
title: "An example of market-clearing assessment"
output:
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{An example of market-clearing assessment}
  %\VignetteEngine{knitr::rmarkdown}
---

This short tutorial gives an example of how one can statistically assess whether a market is in an equilibrium state. The tutorial assumes some familiarity with the concepts and the functionality of the package. The [basic_usage](basic_usage.html) vignette can be helpful in acquiring this familiarity.

```{r, include = FALSE}
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
}
```

## Setup the environment

Load the required libraries.

```{r setup.libraries}
library(diseq)
library(magrittr)
```

Prepare the data. Here, we simply simulate data using a data generating process for a market in equilibrium.

```{r setup.data}
nobs <- 5000
tobs <- 5

alpha_d <- -1.9
beta_d0 <- 4.9
beta_d <- c(2.1, -0.7)
eta_d <- c(3.5, 6.25)

alpha_s <- 2.8
beta_s0 <- 1.2
beta_s <- c(0.65)
eta_s <- c(1.15, 4.2)

sigma_d <- 1
sigma_s <- 1
rho_ds <- 0.5

seed <- 443

eq_data <- simulate_model_data(
  "equilibrium_model", nobs, tobs,
  alpha_d, beta_d0, beta_d, eta_d,
  alpha_s, beta_s0, beta_s, eta_s,
  NA, NA, c(NA),
  sigma_d = sigma_d, sigma_s = sigma_s, rho_ds = rho_ds,
  seed = seed
)
```

## Initialize the model

Prepare the basic parameters for model initialization.

```{r model.parameters}
key_columns <- c("id", "date")
time_column <- c("date")
quantity_column <- "Q"
price_column <- "P"
demand_specification <- paste0(price_column, " + Xd1 + Xd2 + X1 + X2")
supply_specification <- "Xs1 + X1 + X2"
price_specification <- "Xp1"
verbose <- 2
use_correlated_shocks <- TRUE
```

Using the above parameterization, construct the model objects. Here we construct two equilibrium models and four  disequilibrium models. All the models are constructed using the simulated data from a model of market in equilibrium.

```{r model.constructor}
eqmdl <- new(
  "equilibrium_model",
  key_columns,
  quantity_column, price_column,
  demand_specification, paste0(price_column, " + ", supply_specification),
  eq_data[eq_data$date != 1, ],
  use_correlated_shocks = use_correlated_shocks, verbose = verbose
)
bsmdl <- new(
  "diseq_basic",
  key_columns,
  quantity_column, price_column,
  demand_specification, paste0(price_column, " + ", supply_specification),
  eq_data[eq_data$date != 1, ],
  use_correlated_shocks = use_correlated_shocks, verbose = verbose
)
damdl <- new(
  "diseq_deterministic_adjustment",
  key_columns, time_column,
  quantity_column, price_column,
  demand_specification, paste0(price_column, " + ", supply_specification),
  eq_data,
  use_correlated_shocks = use_correlated_shocks, verbose = verbose
)
```

## Estimation 

Set the estimation parameters. 

```{r estimation.parameters}
optimization_method <- "BFGS"
optimization_controls <- list(maxit = 10000, reltol = 1e-8)
```

Estimate the models.

```{r estimation.execution}
eqmdl_reg <- estimate(eqmdl, method = "2SLS")
eqmdl_est <- estimate(eqmdl,
  control = optimization_controls,
  method = optimization_method
)
bsmdl_est <- estimate(bsmdl,
  control = optimization_controls,
  method = optimization_method
)
damdl_est <- estimate(damdl,
  control = optimization_controls,
  method = optimization_method
)
```

## Post estimation analysis 

### Summaries

All the models provide estimates for the simulated data. Even with simulated data, it is difficult to assess which model performs better by examining only the summaries in separation or collectively. 

```{r analysis.summaries}
summary(eqmdl_reg$first_stage_model)
summary(eqmdl_reg$system_model)
bbmle::summary(eqmdl_est)
bbmle::summary(bsmdl_est)
bbmle::summary(damdl_est)
```

### Model selection

The deterministic adjustment model has price dynamics that are analogous to excess demand and estimates one extra parameter. The directional model estimates one parameter less as the model does not have enough equations to identify prices in both demand and supply equations. The estimated parameters are summarized as follows.

```{r analysis.estimates}
sim_coef <- c(
  alpha_d, beta_d0, beta_d, eta_d,
  alpha_s, beta_s0, beta_s, eta_s,
  NA,
  sigma_d, sigma_s,
  rho_ds
)
names(sim_coef) <- names(damdl_est@coef)

dm_inc <- eqmdl_reg$system_model$coefficients[
  grep(
    "demand",
    names(eqmdl_reg$system_model$coefficients)
  )
]
sp_inc <- eqmdl_reg$system_model$coefficients[
  grep(
    "supply",
    names(eqmdl_reg$system_model$coefficients)
  )
]
lm_coef <- c(
  dm_inc[2], dm_inc[-2], sp_inc[2], sp_inc[-2],
  NA,
  NA, NA,
  NA
)

eqmdl_coef <- append(
  eqmdl_est@coef, c(NA),
  after = which(names(eqmdl_est@coef) ==
    get_prefixed_variance_variable(eqmdl@system@demand)) - 1
)

bsmdl_coef <- append(
  bsmdl_est@coef, c(NA),
  after = which(names(bsmdl_est@coef) ==
    get_prefixed_variance_variable(bsmdl@system@demand)) - 1
)

damdl_coef <- damdl_est@coef

comp <- tibble::tibble(
  parameter = names(sim_coef),
  sim = sim_coef, lm = lm_coef, fi = eqmdl_coef,
  bm = bsmdl_coef, da = damdl_coef,
  lmerr = abs(lm_coef - sim_coef), fierr = abs(eqmdl_coef - sim_coef),
  bmerr = abs(bsmdl_coef - sim_coef), daerr = abs(damdl_coef - sim_coef)
)
comp
```

Since we have used simulated data, we can calculate the average absolute error of the parameter estimation for each of the models. In practice, the population values are unknown and this calculation is impossible. 

```{r analysis.averages}
comp_means <- colMeans(comp[, grep("err", colnames(comp))], na.rm = TRUE)
comp_means
```

Moreover, the average absolute error cannot provide an overall estimation assessment as the market models have different parameter spaces. To assess the overall model performance one can instead use an information criterion.

```{r analysis.model.selection}
model_names <- c(
  eqmdl@model_type_string,
  bsmdl@model_type_string, damdl@model_type_string
)
model_obs <- c(
  get_number_of_observations(eqmdl),
  get_number_of_observations(bsmdl),
  get_number_of_observations(damdl)
)
model_errors <- c(
  comp_means["fierr"],
  comp_means["bmerr"],
  comp_means["daerr"]
)
seltbl <- AIC(eqmdl_est, bsmdl_est, damdl_est) %>%
  tibble::add_column(Model = model_names, .before = 1) %>%
  tibble::add_column(Obs. = model_obs, `Mean Error` = model_errors) %>%
  dplyr::rename(D.F. = df) %>%
  dplyr::arrange(AIC)
seltbl
```
